---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(cowplot)
library(broom)
library(here)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. *Is the Curb presence one of the important factors affecting the diameter of trees?*
2. *Is the root_barrier one of the important factors affecting the height of trees?*
3. *What is the relationship between the diameter of a tree and its age?*
4. *Is the distribution of different types of trees related to their geographical location or neighborhood of cultivation?*
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### 1.*Is the Curb presence one of the important factors affecting the diameter of trees?*
**Summarizing:**

I will choose option 1 and compute the *range*, *mean*, *median*, and *variance* of **diameter** across the groups of **curb** from *vancouver_trees*.
```{r}
answer1.21 <- vancouver_trees %>%
  group_by(curb) %>%
  summarise(range =  diff(range(diameter)), mean = mean(diameter),  median = median(diameter), variance = var(diameter))
answer1.21
```
**Graphing:**

I will choose option 8 and create a graph that has at least two geom layers based on the results.
```{r}
p1 <- ggplot(answer1.21) +
  geom_col(aes(x = curb, y = range, fill = curb),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = curb, y = range,label = range)) 
  
p2 <- ggplot(answer1.21) +
  geom_col(aes(x = curb, y = mean, fill = curb),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = curb, y = mean,label = mean)) 

p3 <- ggplot(answer1.21) +
  geom_col(aes(x = curb, y = median, fill = curb),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = curb, y = median,label = median)) 
  

p4 <- ggplot(answer1.21) +
  geom_col(aes(x = curb, y = variance, fill = curb),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = curb, y = variance,label = variance)) 

plot_grid(p1, p2, p3, p4, labels = LETTERS[1:4], ncol = 2)
```

### 2.*Is the root_barrier one of the important factors affecting the height of trees?*
**Summarizing:**
I will choose option 1 and compute the *range*, *mean*, *median*, and *variance* of **height_range_id** across the groups of **root_barrier** from *vancouver_trees*.
```{r}
answer1.22 <- vancouver_trees %>%
  group_by(root_barrier) %>%
  summarise(range =  diff(range(height_range_id)), mean = mean(height_range_id),  median = median(height_range_id), variance = var(height_range_id))
answer1.22
```
**Graphing:**

I will choose option 8 and create a graph that has at least two geom layers based on the results.
```{r}
p1 <- ggplot(answer1.22) +
  geom_col(aes(x = root_barrier, y = range, fill = root_barrier),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = root_barrier, y = range,label = range)) 
  
p2 <- ggplot(answer1.22) +
  geom_col(aes(x = root_barrier, y = mean, fill = root_barrier),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = root_barrier, y = mean,label = mean)) 

p3 <- ggplot(answer1.22) +
  geom_col(aes(x = root_barrier, y = median, fill = root_barrier),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = root_barrier, y = median,label = median)) 
  

p4 <- ggplot(answer1.22) +
  geom_col(aes(x = root_barrier, y = variance, fill = root_barrier),position = position_dodge(),width = 0.4) +
  geom_text(aes(x = root_barrier, y = variance,label = variance)) 

plot_grid(p1, p2, p3, p4, labels = LETTERS[1:4], ncol = 2)
```

### 3.*What is the relationship between the diameter of a tree and its age?*
**Summarizing:**
I will choose option 3 and create a categorical variable with 3 or more groups from an existing numerical variable. In this case, I will create a new categorical variable called diameter_category based on the original value of diameters, which is divided into three groups:

+ 1. low: 0 < diameter <= 5
+ 2. moderate: 5 < diameter <= 15
+ 3. high: diameter > 15
```{r}
answer1.23 <- vancouver_trees %>%
  select(tree_id, diameter) %>%
  mutate(diameter_categorty = factor(case_when(diameter <= 5 ~"low",
                                               diameter <= 15 ~"moderate",
                                               TRUE ~ "high"),
                                     levels = c('low', 'moderate', 'high'))) %>% 
  group_by(diameter_categorty) %>%
  count()
answer1.23
```
From the result, we could see that there are 50919 trees with low diameter_categorty, 54490 trees with moderate diameter_categorty and 50919 trees with high diameter_categorty respectively.

**Graphing:**
I will choose option 7 and make a graph where it makes sense to customize the alpha transparency.In this case, I intend to explore how age would affect trees' diameter.
```{r}
vancouver_trees %>%
  ggplot(aes(date_planted, diameter)) +
  geom_point(alpha = 0.01, size = 1, color = "blue") +
  scale_y_continuous(limits = c(0, 60))
```

Due to the fact that there are so many data points in the plot causing overlapping, it make sense to change alpha transparency which can help reveal the density of points.

### 4.*Is the distribution of different types of trees related to their geographical location or neighborhood of cultivation?*
**Summarizing:**
I will choose option 2 and compute the number of observations for the variable `genre_name`, and explore its potential relationship with geographical location and neighborhood.
```{r}
#compute the number of observations for the variable `genre_name` with different neighbourhood_name
answer1.25 <- vancouver_trees %>%
  group_by(neighbourhood_name, genus_name) %>%
  count()
answer1.25
```
**Graphing:**
I will choose option 7 and make a graph where it makes sense to customize the alpha transparency. In this section, I will make a plot to show the geographical distribution of the top 10 planted trees.
```{r}
#count and sort genus_name
answer1.24 <- vancouver_trees %>%
  group_by(genus_name) %>%
  count() %>%
  arrange(-n)
answer1.25 <- answer1.24[1:10,]
answer1.25
```
```{r}
answer1.26 <- vancouver_trees %>%
  filter(genus_name %in% answer1.25$genus_name) %>%
  ggplot(aes(x = longitude, y = latitude, color = genus_name)) +
  geom_point(alpha = 0.1, size = 1)
answer1.26
```

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
### 1.*Is the Curb presence one of the important factors affecting the diameter of trees?*
Based on the results of 1.2, we could conclude that Curb presence is one of the important factors that affect the diameter of trees. From the plotted graph, we could find that compared with trees with no curb, the diameter of trees with curbs has higher range, but their mean and median value are lower. Besides, the variance of the diameter of curbed trees are lower than those without curbs. However, we have not explore if the presence of curb is related to genus, or if the effect of curb is related to genus. I will revise this question and continue further study.

### 2.*Is the root_barrier one of the important factors affecting the height of trees?*
Based on the results of 1.2, we could conclude that root_barrier is one of the important factors that affect the height of trees. From the plotted graph, we could find that compared with trees with no root_barrier, the height of trees with root_barrier has lower mean and median value. Besides, the variance and range of the height of trees with root_barrier are also lower than others. However, I also notice that the height_range_id is a approximation of trees' height, which is not precise enough to derive convinced results.

### 3.*What is the relationship between the diameter of a tree and its age?*
Based on the results of 1.2, we could find that older trees tend to have higher diameter than young trees, as shown in the plotted graph. However, given that we have not consider other factors affecting the diameters such genus and geographical location, further research need to be implemented to confirm our assumption. However,apart from ages, the diameter of trees might also affected by trees' genus. Further proof need to be done to verify our conclusion.

### 4.*Is the distribution of different types of trees related to their geographical location or neighborhood of cultivation?*
Based on the results of 1.2, we still could not find any obvious laws applied to the geographical distribution of different genres of trees. I try to plot the top 10 planted trees and make them shown as a map, but its hard to  distinguish its distribution pattern with the naked eye. Further research methods and algorithms might applied to help solve this question.
<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
```{r}
glimpse(vancouver_trees)
```
Due to the fact that the dataset *vancouver_trees* contains more than 8 variables, I will pick 8 out of them and do further explanation.
```{r}
answer2.11 <- vancouver_trees %>%
  reframe(tree_id, civic_number, std_street, genus_name, species_name, common_name, assigned, root_barrier)
answer2.11

```
As shown above, each row is an observation, each column is a variable and each cell is a value, so I think my data is tidy.


<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
**Before Untidy**

```{r}
head(vancouver_trees)
```
**Process of Untidy**

In order to untidy my data, I try to merge or combine relative columns. In this section, I will merge `genus_name` and `species_name` to a single column.
```{r}
answer2.12 <- answer2.11 %>% 
  pivot_longer(cols = c(genus_name, species_name), names_to = "species_or_genus", values_to = "value")
answer2.12
```

**Tidy Back**
```{r}
answer2.13 <- answer2.12 %>% 
  pivot_wider(names_from = species_or_genus, values_from = value)
answer2.13
```

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

### 1. *Is the Curb presence one of the important factors affecting the diameter of trees, regardless of genus?*
### 2. *What is the relationship between the diameter of a tree and its age? Does genus affect trees' diameter?*

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->
The reason that I finally choose these two questions to continue analysis is that I found them more related to the course content, and also more practical to explore. Through my consideration, the `height_range_id` could not represent the height of trees precisely, so I abandon original research question 2. Besides, regarding original research question 4, I have not worked out an effective method to analyse the potential relationship between trees' geographical location and their genus.
<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->
### 1. *Is the Curb presence one of the important factors affecting the diameter of trees, regardless of genus?*
In this case, I will use select(), summarise(), pivot_wider() and filter(). First, we want to compare the diameter of trees of the same genus with and without curbs.
```{r}
answer2.31 <- vancouver_trees %>%
  select(genus_name, curb, diameter) %>% 
  group_by(curb, genus_name) %>%
  summarise(Mean = mean(diameter)) %>%
  pivot_wider(names_from = curb, values_from = Mean)
answer2.31
```
Then I try to check how many genus tend to have higher diameter with curb.
```{r}
answer2.32 <- answer2.31 %>% 
  count(Y > N)
answer2.32
```

From the result we can find that, although for all trees, the presence of curb would decrease the mean value of diameter, trees with different genus tend to be affected differently. Apart from 19 NA genus, 39 genus have lower mean value of diameter, while other 39 genus are opposite. 

### 2. *What is the relationship between the diameter of a tree and its age? Does genus affect trees' diameter?*
In this case,  I will use group_by(), summarise(), mean(), median() variance() and range() to explore potential relationship between trees' genus and diameter.
```{r}
answer2.33 <- vancouver_trees %>%
  group_by(genus_name) %>%
  summarise(mean(diameter), median(diameter), var(diameter), diff(range(diameter)))
answer2.33
```

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: What is the relationship between the diameter of a tree and its age?

**Variable of interest**: diameter

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->
```{r}
answer3.11 <- lm(diameter ~ date_planted, vancouver_trees)
answer3.11
```

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->
In this section, I will use functions within `broom` package to obtain coefficient values, standard errors, t-statistics, p-values, etc.
```{r}
answer3.12 <- tidy(answer3.11)
answer3.12
```
```{r}
answer3.13 <- augment(answer3.11)
answer3.13
```
```{r}
answer3.14 <- glance(answer3.11)
answer3.14
```

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->
I will choose answer1.25 to be the summary table.
```{r}
answer1.25
```
```{r}
summary_table <- answer1.25
if(!dir.exists(here("output"))){
  dir.create(here("output"))
}
here()
write_csv(summary_table, here("output", "summary_table.csv"))
```
<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->
```{r}
answer3.11
```

```{r}
#Write your model object from Task 3 to an R binary file
 model <- answer3.11
if(!dir.exists(here("output"))){
  dir.create(here("output"))
}
here()
saveRDS(model, here("output", "model.rds"))
```
```{r}
#load the model again
readRDS(here("output", "model.rds"))
```

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.